The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2021-01-15 Rmd/trajectories_manuscript.Rmd
Separate vocab type

author: 
  - name          : "Gonzalo García-Castro"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain"
    email         : "gonzalo.garciadecastro@upf.edu"
  - name          : "Daniela Avila-Varela"
    affiliation   : "1"
  - name          : "Núria Sebastian-Galles"
    affiliation   : "1"

@@

# load packages
library(papaja)
library(knitr)
library(magrittr)
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
library(gt)
library(data.table)
library(here)

@@

The structure of the lexicon also impacts the order of acquisition of new words. Words that share a high degree of phonological and/or semantic overlap with words already acquired are more likely to be acquired next [@hills2009]. @borosky2016 and @borovsky2018 showed that the connectivity of a given semantic category (i.e., animals) in a child's lexicon predicts a better performance in a disambiguation task, where participants are presented with a novel label in the presence of a familiar and a novel object: infants show stronger looking preference for the novel object if it belonged to a category for which many words had already been acquired. This points to the structure of the lexicon facilitating the strategies children engage during word learning.

@@

When compared to their monolingual peers, they know fewer words only when just on language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When both languages are taken into account, bilingual children  seem to know, at least, as many words as monolinguals do [@ben-zeev_influence_1977; @doyle1977some; @rosenblum1983word; @fernandez1992bilingual; @pearson_patterns_1994; @bialystok_receptive_2010; @hoff_dual_2012; @core_total_2013; @hoff_expressive_2014; @blom_cross-language_2019; but see @pearson_lexical_1993; @houwer_bilingualmonolingual_2014].

@floccia_i_2018, gathered parental vocabulary estimates [@hamilton2000infant] from a large sample of 24-month-old toddlers learning English and an additional language (from a pool of 13 linguistically diverse languages). They computed the average phonological similarity between translation equivalents (TEs from now on; e.g.,  *table* in English and *tafel* in Dutch) for each pair of languages, and found a positive association between this measure of phonological similarity and participants' productive vocabulary sizes in their additional language. Using a similar measure of lexical similarity, @blom_cross-language_2019 extended these results to bilingual children aged three to 10, and reported a positive association between the lexical distance and comprehensive vocabulary size: bilinguals learning lexically close languages showed similar vocabulary sizes to those of monolinguals, but those learning the more distant languages showed lower vocabulary sizes.

These results suggest that linguistic distance between languages plays an important role during bilingual lexical acquisition, and that this effect operates within TEs. What mechanisms underlie this effect, and why they seem to operate differently across comprehension and production, or across the dominant and the non-dominant language, remain open issues. @floccia_i_2018 pointed to parallel activation as a candidate mechanism behind this facilitation effect. The parallel activation principle suggests that bilinguals activate lexical representations in a language non-selective way: during the comprehension or production of a word in one language (e.g., *cat*, in English) its corresponding translation in the other language is activated too [e.g., *gato*, in Spanish. There is a vast body of evidence supporting language non-selective lexical access in both adults [e.g., @dell1991mediated; @spivey1999cross; @costa2000lexical; @dijkstra_bilingual_2005; @thierry_brain_2007; @hoshino_cognate_2008; @yudes_cognate_2010; @kroll_multilingual_2013; @singh2014one; @von_holzen_impact_2019; @bobb_co-activation_2020; see @costa2005facilitatory for review], and children [e.g. @von_holzen_language_2012; @poulin2013lexical; @jardak2019labels; @bosma2020cognate]. A critical implication of parallel activation is that the word form (i.e., phonology, orthography, etc.) of both translation equivalents are available to each other, impacting the comprehension or production dynamic of any of them.

@@

Under this account, cognates (i.e., phonologically similar TEs) would be acquired earlier than non-cognates, but this effect would only play a role once one of the members of the TE has been acquired. The evidence supporting an earlier age of acquisition of cognates relative to non-cognates if sparse. @schelletter2002effect's reported a longitudinal case-study of a German/English child who produced cognate TEs earlier than non-cognate TE. @bosch_first_2014, showed converging results from a sample of 48 monolingual and bilingual infants aged 24 months. The low number of words included in the analyses of both studies, and the lack of adjustment for lexical frequency, limits severely the strength of any conclusions concerning the difference in age of acquisition of cognates and non-cognates. On the other hand @floccia_i_2018 worked with aggregated estimates of participants' vocabulary size, losing item-level information that could provide information about change in age of acquisition of TEs associated with their degree of phonological similarity. Therefore, their results can only be considered as compatible with cognateness playing a role on age of acquisition. In this study, we aimed at overcoming these pitfalls by using unaggregated data from individual responses, testing the role of cognateness explicitly and adjusting for lexical frequency.

@@

We then fit increasingly simpler models dropping predictors one by one in the following order: (1) `dominance:bilingualism:cognate`, (2) `bilingualism:cognate`, (3) `dominance:cognate:`, (4) `cognate`, (5) `dominance:bilingualism`, (6) `bilingualism`, (7) `dominance`, (8) `frequency`, and (9) `age`. We compared all models using leave-one-out cross-validation (LOO-CV) [@vehtari_practical_2017] and widely applicable information criterion (WAIC) [@watanabe_asymptotic_2010]. We then explored the posterior distribution of each parameter in the model that fitted the data the best, computing the 95% credible intervals and testing whether this interval excluded 0. We performed follow-up tests on interactions whose 95% credible interval excluded 0 by comparing the 95% credible interval of their estimated marginal means using the `emmeans` package [@lenth_emmeans_2020]. Data processing and visualisation was done in R [@rcoreteam_r_2020] using the `tidyverse` family of packages [@wickham_welcome_2019], and posterior samples were extracted using the `tidybayes` R package [@kay_tidybayes_2020].

@@

>>>>

author: 
  - name          : "Gonzalo Garcia-Castro"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain"
    email         : "gonzalo.garciadecastro@upf.edu"
  - name          : "Daniela Avila-Varela"
    affiliation   : "1"
  - name          : "Nuria Sebastian-Galles"
    affiliation   : "1"

@@

# load packages
library(papaja)
library(knitr)
library(magrittr)
library(tidyverse)
library(readxl)
library(lubridate)
library(brms)
library(tidybayes)
library(janitor)
library(gt)
library(data.table)
library(here)

source(here("R", "utils.R"))
```

@@

The structure of the lexicon also impacts the order of acquisition of new words. Words that share a high degree of phonological and/or semantic overlap with words already acquired are more likely to be acquired next [@hills_longitudinal_2009]. @borosky2016 and @borovsky2018 showed that the connectivity of a given semantic category (i.e., animals) in a child's lexicon predicts a better performance in a disambiguation task, where participants are presented with a novel label in the presence of a familiar and a novel object: infants show stronger looking preference for the novel object if it belonged to a category for which many words had already been acquired. This points to the structure of the lexicon facilitating the strategies children engage during word learning.

@@

When compared to their monolingual peers, they know fewer words only when just on language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When both languages are taken into account, bilingual children  seem to know, at least, as many words as monolinguals do [@ben-zeev_influence_1977; @doyle_some_1977; @rosenblum_word_1983; @fernandez_bilingual_1992; @pearson_patterns_1994; @bialystok_receptive_2010; @hoff_dual_2012; @core_total_2013; @hoff_expressive_2014; @blom_cross-language_2019; but see @pearson_lexical_1993; @houwer_bilingualmonolingual_2014].

@floccia_i_2018, gathered parental vocabulary estimates [@hamilton_infant_2000] from a large sample of 24-month-old toddlers learning English and an additional language (from a pool of 13 linguistically diverse languages). They computed the average phonological similarity between translation equivalents (TEs from now on; e.g.,  *table* in English and *tafel* in Dutch) for each pair of languages, and found a positive association between this measure of phonological similarity and participants' productive vocabulary sizes in their additional language. Using a similar measure of lexical similarity, @blom_cross-language_2019 extended these results to bilingual children aged three to 10, and reported a positive association between the lexical distance and comprehensive vocabulary size: bilinguals learning lexically close languages showed similar vocabulary sizes to those of monolinguals, but those learning the more distant languages showed lower vocabulary sizes.

These results suggest that linguistic distance between languages plays an important role during bilingual lexical acquisition, and that this effect operates within TEs. What mechanisms underlie this effect, and why they seem to operate differently across comprehension and production, or across the dominant and the non-dominant language, remain open issues. @floccia_i_2018 pointed to parallel activation as a candidate mechanism behind this facilitation effect. The parallel activation principle suggests that bilinguals activate lexical representations in a language non-selective way: during the comprehension or production of a word in one language (e.g., *cat*, in English) its corresponding translation in the other language is activated too [e.g., *gato*, in Spanish. There is a vast body of evidence supporting language non-selective lexical access in both adults [e.g., @dell_mediated_1991; @spivey_cross_1999; @costa_lexical_2000; @dijkstra_bilingual_2005; @thierry_brain_2007; @hoshino_cognate_2008; @yudes_cognate_2010; @kroll_multilingual_2013; @singh_one_2014; @von_holzen_impact_2019; @bobb_co-activation_2020; see @costa_facilitatory_2005 for review], and children [e.g. @von_holzen_language_2012; @poulin_lexical_2013; @jardak_labels_2019; @bosma_cognate_2020]. A critical implication of parallel activation is that the word form (i.e., phonology, orthography, etc.) of both translation equivalents are available to each other, impacting the comprehension or production dynamic of any of them.

@@

Under this account, cognates (i.e., phonologically similar TEs) would be acquired earlier than non-cognates, but this effect would only play a role once one of the members of the TE has been acquired. The evidence supporting an earlier age of acquisition of cognates relative to non-cognates if sparse. @schelletter_effect_2002 reported a longitudinal case-study of a German/English child who produced cognate TEs earlier than non-cognate TE. @bosch_first_2014, showed converging results from a sample of 48 monolingual and bilingual infants aged 24 months. The low number of words included in the analyses of both studies, and the lack of adjustment for lexical frequency, limits severely the strength of any conclusions concerning the difference in age of acquisition of cognates and non-cognates. On the other hand @floccia_i_2018 worked with aggregated estimates of participants' vocabulary size, losing item-level information that could provide information about change in age of acquisition of TEs associated with their degree of phonological similarity. Therefore, their results can only be considered as compatible with cognateness playing a role on age of acquisition. In this study, we aimed at overcoming these pitfalls by using unaggregated data from individual responses, testing the role of cognateness explicitly and adjusting for lexical frequency.

@@

We fit increasingly simpler models dropping predictors one by one in the following order: (1) `dominance:bilingualism:cognate`, (2) `bilingualism:cognate`, (3) `dominance:cognate:`, (4) `cognate`, (5) `dominance:bilingualism`, (6) `bilingualism`, (7) `dominance`, (8) `frequency`, and (9) `age`. We compared all models using leave-one-out cross-validation (LOO-CV) [@vehtari_practical_2017]^[Due to the high computational cost associated with a large dataset, we used a sub-sampling approach for performing Bayesian LOO with 1,000 samples [@magnusson_leave_2019]]. We then explored the posterior distribution of each parameter in the model that fitted the data the best, computing the 95% credible intervals and testing whether this interval excluded 0. We performed follow-up tests on interactions whose 95% credible interval excluded 0 by comparing the 95% credible interval of their estimated marginal means using the `emmeans` package [@lenth_emmeans_2020]. Data processing and visualisation was done in R [@rcoreteam_r_2020] using the `tidyverse` family of packages [@wickham_welcome_2019], and posterior samples were extracted using the `tidybayes` R package [@kay_tidybayes_2020].

@@

```{r results}
fit_9 <- readRDS(here("Results", "fit_9.rds"))
loos <- readRDS(here("Results", "loos.rds"))

Table 2 summarises shows the outcome of the LOO-CV model comparison. The extended model was most favoured by the data.

```{r loos, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

as.data.frame(loos$diffs) %>% 
  mutate(
    model = md(c(
      "~ 1 (intercept only)",
      "... + age",
      "... + frequency",
      "... + dominance",
      "... + bilingualism",
      "... + dominance:bilingualism",
      "... + cognate",
      "... + dominance:cognate",
      "... + bilingualism:cognate",
      "... + dominance:bilingualism:cognate"
    ))) %>% 
  select(model, elpd_loo, se_elpd_loo, p_loo, elpd_diff, se_diff) %>% 
  gt() %>% 
  tab_spanner(
    label = md("**Comparison**"),
    columns = c("elpd_diff", "se_diff")
  ) %>% 
  cols_label(
    model = md("**Model**"),
    elpd_loo = md("**ELPD**"),
    se_elpd_loo = md("**SE**"),
    p_loo = md("**p**"),
    elpd_diff = md("*LOO<sub>diff</sup>*"),
    se_diff = md("*SE<sub>diff</sup>*")
  ) %>% 
  fmt_number(2:5) %>% 
  tab_footnote(
    footnote = md("*ELPD*: theoretical expected log pointwise predictive density for a new dataset, as estimated using Bayesian LOO-CV"),
    locations = cells_column_labels(columns = vars("elpd_loo"))
  ) %>% 
  tab_footnote(
    footnote = md("*SE*: standard error of the ELPD"),
    locations = cells_column_labels(columns = vars("se_elpd_loo"))
  ) %>% 
  tab_footnote(
    footnote = md("*p*: effective number of parameters"),
    locations = cells_column_labels(columns = vars("p_loo"))
  ) %>% 
  tab_options(footnotes.marks = "letters")

Table 3 shows the mean and 95% CrI of each regression coefficient of the best fitting model.

```{r coefs, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

cbind(as.data.frame(exp(fixef(fit_9))),
      as.data.frame(inv_logit_scaled(fixef(fit_9)))) %>% 
  rownames_to_column("coefficient") %>% 
  mutate(coefficient = c(
    "Intercept 1 (Understands)",
    "Intercept 2 (Understands & Says)",
    "Age",
    "Frequency",
    "Dominance",
    "Bilingualism",
    "Cognate",
    "Dominance * Bilingualism",
    "Dominance * Cognate",
    "Bilingualism * Cognate",
    "Dominance * Bilingualism * Cognate"
  )) %>% 
  clean_names() %>% 
  gt() %>% 
  tab_spanner(label = md("**Odds Ratio**"), columns = 2:5) %>% 
  tab_spanner(label = md("**Probability**"), columns = 6:8) %>% 
  fmt_number(2:9) %>% 
  cols_merge_range(col_begin = "q2_5", col_end = "q97_5") %>% 
  cols_merge_range(col_begin = "q2_5_1", col_end = "q97_5_1") %>% 
  cols_label(
    coefficient = md("**Coefficient**"),
    estimate = md("*Mean*"),
    est_error = md("*SE*"),
    q2_5 = md("*95% CrI*"),
    estimate_1 = md("*Mean*"),
    est_error_1 = md("*SE*"),
    q2_5_1 = md("*95% CrI*"),
  ) %>% 
  tab_footnote(
    footnote = "SE: Standard Error of the mean",
    locations = cells_column_labels(columns = vars("est_error"))
  ) %>% 
  tab_footnote(
    footnote = "95% CrI: 95% credible interval",
    locations = cells_column_labels(columns = vars("q2_5"))
  )

```{r marginalmeans, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=6.5, fig.height=6.5, out.width="20cm", fig.cap="Mean posterior probability and 95\\% CrI of responses, conditional to Dominance, Bilingualism, and Cognate."}

# summarise data
proportion <- responses %>% 
  mutate(
    understands = response %in% 2,
    says = response %in% 3, 
    frequency = cut_quantiles(frequency),
    age = round(age)
  ) %>%
  select(-doe) %>% 
  pivot_longer(c(understands, says), names_to = ".category", values_to = ".value") %>% 
  mutate(
    .category = str_to_sentence(.category),
    .value = as.numeric(.value)
  ) %>% 
  group_by(age, cognate, item_dominance, .category) %>%
  summarise(
    .value = sum(.value),
    n = n(),
    .groups = "drop"
  ) %>% 
  rowwise() %>% 
  mutate(.value = prop_adj(.value, n)) %>% 
  ungroup()

# get 20 posterior draws
# values to get posterior draws for (set frequency at mean = 0)
nd <- expand_grid(
  age = seq(min(responses$age), max(responses$age), 1),
  item_dominance = c("L1", "L2"),
  doe = c(-1, 0, 1),
  frequency = 0,
  cognate = c("Cognate", "Non-Cognate")
)

post_preds <- add_fitted_draws(nd, fit_9, n = 20) %>% 
  mutate(doe = doe %>%
           as.character() %>%  
           str_replace_all(c("0" = "Mean","1" = "1 SD")) %>% 
           paste0("DOE = ", .) %>% 
           factor(levels = c("DOE = -1 SD", "DOE = Mean","DOE = 1 SD"), ordered = TRUE)) %>% 
  mutate(
    .category = case_when(
      .category==1 ~ "None",
      .category==2 ~ "Understands",
      .category==3 ~ "Says"
    ),
    .category = factor(.category, levels = c("None", "Understands", "Says"), ordered = TRUE)
  )

# visualise posterior predictions
post_preds %>% 
  filter(.category!="None") %>% 
  ggplot(aes(
    age, .value,
    colour = interaction(item_dominance, cognate, sep = " - "),
    fill = interaction(item_dominance, cognate, sep = " - ")
  )) +
  facet_grid(doe~.category) +
  #geom_line(aes(group = interaction(item_dominance, cognate, doe, .draw)), size = 0.4) +
  stat_lineribbon(.width = 0.95, colour = NA, alpha = 0.5) +
  stat_summary(fun = "median", geom = "line", size = 0.75) +
  geom_point(data = proportion, alpha = 0.5) +
  labs(x = "Age (months)", y = "P(Y|X)") +
  guides(colour = guide_legend(ncol = 2)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_custom() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.direction = "horizontal"
  )

The age at which the average item changed
